---
title: "Affinity Release Model"
author: "Edgardo Rivera-Delgado"
date: '2018-03-01'
output:
  html_document:
    df_print: paged
  html_notebook: default
  word_document: default
csl: biomaterials.csl
bibliography: QSPR.bib
---

# Equations for Drug Release

Release from affinity polymer disks can be modeled as described by Andrew Fu. [@Fu:2011cj]. 



In the next set of equations the continuous equation will be given followed by the
discretized version used to solve it with the method of lines. 

The spatial derivative is discretized as given

$$ \Delta = \frac{1}{N}, z_{i}^{*} = i\Delta \\
z^{*} = \frac{z}{\delta}\\
(i = 0,1,2,...,N) \\
z = \frac{\Sigma (0,1,2,...,N) *\delta} {N}??$$

The rate of binding within the hydrogel is given by

$$ R_b = \bigg[\frac{k_1C_0} {k_2}\bigg] C_L\bigg(\bigg[\frac{C_T}{C_0}\bigg]-C_{L-C}\bigg)-C_{L-C} \\
R_{b,i}= P_1C_{L,i}\bigg(P_3-C_{L-C,i}\bigg)-C_{L-C,i}$$ 

Free drug in hydrogel varies according to

$$\frac {\partial C_L}{\partial \tau} = \bigg[ \frac {D}{k_2 \delta^2} \bigg ]\frac{\partial^2C_L}{\partial z^2}-R_b  \\
\frac{dC_{L,i}}{d\tau} = P_2 \frac{C_{L,i+1}+2C_{L,i}+C_{L,i-1}}{\Delta^2}-R_{b,i} \\
(i = 1,2, ..., N-1 )$$


Drug to cyclodextrin complex varies according to

$$\frac {\partial C_{L-C}} {\partial \tau} = R_b \\
\frac{\partial C_{L-C}} {\partial \tau} = R_{b,i} \\
(i = 1,2, ..., N-1 )$$

With dimensionless boundary conditions at the center of the hydrogel given by

$$z = 0 , \frac {\partial C_L}{\partial z} = 0 \\
i = 0: \frac{\partial C_{L,0}}{\partial z} = 0 $$
For the release phase the dimensionless initial conditions are given by

$$ \tau = 0, C_L = \frac {C_L^{eq}} {C_0}, C_{L-C} = \frac {C_L^{eq}}{C_0} $$
which are the same initial conditions for the discretized version.

Conditions at the boundary of the hydrogel and drug release solution 

$$\tau = 0, z = 1, C_L = 0 \\
\tau = 0, i = 1: C_{L,N}=0$$

The rate of release is given by the following equation
$$\frac {dM_R} {d \tau} = - \frac {1}{2} \bigg[\frac{D} {k_2 \delta^2}\bigg] \frac {\partial C_l}{\partial z} \\ 
\frac {dM_R} {d \tau}=-\frac{1}{2}P_2\frac{C_{L,N}-C_{L,N-1}}{\Delta}$$


And with initial conditions given by:

$$M_R = 0, \tau = 0$$

$$ P_1 = \bigg[\frac{K_1*C_0}{K_2}\bigg], P_2 = \bigg[\frac{D}{K_2\delta^2}\bigg], P_3 = \bigg[\frac{C_T}{C_0}\bigg]$$
```{r}
affinity <- function(t, state , parms, ...){

  
# Assigning parameter values    
p1 = k1*C_o/k2
p2 = 0.933
p3 = C_T/C_o


# Extracting values passed onto the function through the state parameter
  LIGAND <- state[1:N]
  COMPLEX <- state[(N+1):(2*N)]
  RELEASE <- state[2*N+1]

# Initializing the derivatives vectors
  dLIGAND <- rep(0, times = N)
  dCOMPLEX <- rep(0, times = N)
  dRELEASE <- 0

# Calculating the rate of change of the state variables

Rb <- p1*LIGAND*(p3 - COMPLEX) - COMPLEX

dCOMPLEX <- Rb

dLIGAND[1] <- p2 * (LIGAND[2] - LIGAND[1])/(delz^2)- Rb[1]
dLIGAND[N-1] <-  p2 * (-2*LIGAND[N-1]+LIGAND[N-2])/(delz^2) - Rb[N-1]
for(i in 2:(N-2)){
dLIGAND[i] <- p2*(LIGAND[i + 1]- 2*LIGAND[i] + LIGAND[i-1])/(delz^2) - Rb[i]}


  
dRELEASE <- -(1/2)*p2*(LIGAND[N]-LIGAND[(N-1)])/(delz^2)


dudt <- rep(0, (2*(N)+1))
dudt <- c(dLIGAND, dCOMPLEX, dRELEASE)

  ncall <<- ncall+1 
return(list(c(dudt))) 
             
  }
```

This model translates to deSolve in the following way
```{r}
library(deSolve)

# Parameters
N   = 50 # Number of boxes (dimensionless)
M_l = .015 # Initial Loading concentration in millimoles.
bcd_MW = 1134.98 # (mg/millimol)
bcd_disk_mass = 15 # (mg)
V_h = 0.0785 # hydrogel volume in (ml)
k2  = 36.9 # h^-1
k1  = 33.7*k2*1000 # mM^-1*h^-1
K = k1/k2 # mM^-1

#(0.7*bcd_disk_mass/bcd_MW)/V_h # mmol/ml == mM
# C_T = C_c + C_LC, which means that the total amount of cd is fixed and exists 
# in either the free of ligand (C_c) or bound (C_LC )state

C_o = M_l # Loading concentration  (mg/ml) C_o = C_LC_eq+C_L_eq = M_l/V_h. Should be below C_T

# Total CD free to bind to drug
# bound_fraction <- 
C_c = 0.09*C_o    # A fraction all Drug Bound
# C_c = C_T - C_o * 0 # ALL CD FREE
# C_c = C_T - C_o * bound_fraction 
# C_c = C_o
C_T = 1.75*(C_c+C_o) #0.1178505
(L_eq = ((C_o)/(1+K*C_c))) # Ligand Equilibrium
(complex_eq = (C_o-((C_o)/(1+K*C_c)))) # 
# From Alison
#complex_eq = C_T - C_c
#L_eq = C_o - complex_eq

thickness = 0.05 # cm
D   = 0.933*k2*(thickness^2) # Diffusion coefficient
delz = 1/N # dimensionless



#LIGAND = rep(L_eq/(C_o*(N-1)), times = N-1)
LIGAND = rep(L_eq/(C_o*(N-1)), times = N)
LIGAND[N] <- 0
COMPLEX = rep(complex_eq/(C_o*N), times = N)
#COMPLEX = rep(complex_eq, times = N)
RELEASE <- 0

state <- c(LIGAND = LIGAND,
           COMPLEX = COMPLEX,
           RELEASE = RELEASE)

```


```{r}
ncall <- 0
times <- seq(0, 700*k2, by = 1)
out <- ode(func = affinity,
           times = times, 
           y = state, parms = NULL, method = "daspk")
```

Now checking the concentrations within the hydrogel below. We can plot the ligand
and complex concentration side by side. You can see that the concentration of 
complex (yellowish region) is much higher relative to the ligand which appears
pitch blue for the duration of the experiment. Does this means that the ligand
is instantaneously depleted from the matrix?

```{r}
library(tidyverse)

tidy_out <- as.data.frame(out) %>% 
  select(-RELEASE) %>%
  gather(., key = "species", value = "concentration", -time) %>%
  separate(col = species, 
           into = c("species", "z"), 
           sep = "(?<=[:alpha:])(?=[:digit:])") %>%
  mutate(z = as.numeric(z)) %>%
  mutate(time = time/k2) # time in hrs
```

```{r}
library(viridis)
library(ggthemes)
tidy_out %>%
  filter(species %in% c("LIGAND", "COMPLEX")) %>%
  ggplot(., aes(x= time, y = z))+
  geom_raster(aes(fill = `concentration`),
              hjust = 0.5,
              vjust= 0.5, 
              interpolate = F)+
  facet_wrap(~species)+
  scale_fill_viridis()+
  theme_tufte()
```

In reality the ligand is not depleted from the matrix but present in very low 
concentrations when compared to the complex. If we plot just the ligand 
concentration we get a different picture as shown below.
```{r}
tidy_out %>%
  filter(species %in% c("LIGAND"), time > 3, time < 700/k2) %>% 
  ggplot(., aes(x= time, y = z))+
  geom_raster(aes(fill = `concentration`),
              hjust = 0.5,
              vjust= 0.5, 
              interpolate = F)+
  scale_fill_viridis()+
  theme_tufte()
```
```{r}
tidy_out %>%
  filter(time > 5,  time < 700/k2, species %in% c("LIGAND")) %>%
  ggplot(., aes(x = z, y = concentration, color = time))+
  facet_wrap(~species)+
  geom_point()
```




```{r}
out %>% 
  as.data.frame() %>%
  select(time, RELEASE) %>%
  ggplot(., aes(x = time/k2, y = RELEASE))+
  geom_point()+
  theme_gray()
```

This following piece of code should be a reality check. It gives the last two
layers of the ligand in the gel that should give a sense to what's happening


```{r}
tidy_out %>% 
  filter(species == "LIGAND", z %in% c((N-1):N)) %>% 
  group_by(z) %>% 
  mutate(ligand =  concentration) %>% 
  ggplot(., aes(x = time, y = ligand, color = as.factor(z)))+
  geom_point()
```


```{r}
tidy_out %>%
  filter(species == "LIGAND", z %in% c(1:N)) %>% 
  group_by(z) %>% 
  ggplot(., aes(x = time, y = concentration, color = z))+
  geom_point()
```
```{r}
tidy_out %>%
  filter(species == "COMPLEX") %>%
  group_by(time) %>%
  summarize(complex = sum(concentration)) %>% 
  ggplot(., aes(x = time, y = complex))+
  geom_point()
```
```{r}
tidy_out %>%
  filter(species == "LIGAND") %>%
  group_by(time) %>%
  summarize(ligand = sum(concentration)) %>% 
  ggplot(., aes(x = time, y = ligand))+
  geom_point()
```
```{r}
tidy_out %>%
  filter(species == "LIGAND") %>%
  group_by(time) %>%
  summarise(sum = sum(concentration)) %>% 
  ungroup() %>%
  mutate(cumulative = cumsum(sum))%>%
  ggplot(., aes(x = time, y = cumulative))+
  geom_point()
```


```{r}
 ligand <- tidy_out %>%
  filter(species == "LIGAND") %>%
  group_by(time) %>%
  summarize(ligand = sum(concentration))
complex <- tidy_out %>%
  filter(species == "COMPLEX") %>%
  group_by(time) %>%
  summarize(complex = sum(concentration))
release <- out %>%
  as.data.frame() %>%
  select(time, RELEASE) %>%
  mutate(time = time / k2)
```

```{r}
inner_join(ligand, complex, by = "time") %>%
  inner_join(release, by = "time") %>%
  as_tibble()%>%
  filter(time > 300) %>%
  mutate(conservation = complex+2*RELEASE) %>%
  ggplot(., aes(x = time, y = conservation))+
  geom_point()+
  geom_point(aes(x = time, y = 2*RELEASE), color = "red")
  # geom_point(aes(x = time, y = cumsum(ligand)), color = "blue")+
  #geom_point(aes(x = time, y = complex), color = "purple")
```
# Possible causes for the lack of mass conservation
  + (1/N)*2 
    + A fractional volume (1/N) of the polymer is sampled and transport assumed to happen through one of the faces(2) but not explicitly stated in the paper. (Done)
    + In such case the reader should be compentent enough to apply it themselves.
  + RR is the actual release through the polymer face and scaling the concentration to the volume of the gel is a flawed reasoning. (Remove the N and N-1 scaling and pass onto another file)
    + Unlikely $C_o = \frac {M_L}{V_h} = C_L_eq + C_L-C_Eq$ and vulic also normalizes total host and ligand concentration to 1
  + The RR is expressed as a first derivative when it should be a second dervative.  
    + This could mean that Fu's paper is incorrect in its approximation as a first derivative. (Ask Saidel or Fu) 
  + Alteratively it could be a problem of numerical approximation. (Email the modeling team) 
    + Explore other solver methods(ode23[x], ode45[!], daspk[!], etc) (Write onto another file)
    + The approximation may just be unstable at the very beginning of the simulation 
  + The system of equations may be wrongly formulated 
    + Complex may be escaping but not ligand
      + Rewrite the function without Rb
    + In the discretization matlab starts at 0 and R starts at index 1 which makes it possible that I am incorrectly translating the system of equations between languages.(Read matlab documentation for ode15s)
    + Alternatively, I could be including an extra index for i = 1 and i = 50 (In Progress)
    + Implement in ode.1D (After previous point is finished)
    + Like Alison, remove RR from the ode and solve separately.
