---
title: "Affinity Release Model"
author: "Edgardo Rivera-Delgado"
output: html_notebook
date: 2018-03-01
bibliography: QSPR.bib
csl: biomaterials.csl
---

# Equations for Drug Release

Release from affinity polymer disks can be modeled as described by Andrew Fu. [@Fu:2011cj]. 


In the next set of equations the continuous equation will be given followed by the
discretized version used to solve it with the method of lines. 

The spatial derivative is discretized as given

$$ \Delta = \frac{1}{N}, z_{i} = i\Delta \\
(i = 0,1,2,...,N)$$

The rate of binding within the hydrogel is given by

$$ R_b = \bigg[\frac{k_1C_0} {k_2}\bigg] C_L\bigg(\bigg[\frac{C_T}{C_0}\bigg]-C_{L-C}\bigg)-C_{L-C} \\
R_{b,i}= P_1C_{L,i}\bigg(P_3-C_{L-C,i}\bigg)-C_{L-C,i}$$ 

Free drug in hydrogel varies according to

$$\frac {\partial C_L}{\partial \tau} = \bigg[ \frac {D}{k_2 \delta^2} \bigg ]\frac{\partial^2C_L}{\partial z^2}-R_b  \\
\frac{dC_{L,i}}{d\tau} = P_2 \frac{C_{L,i+1}+2C_{L,i}+C_{L,i-1}}{\Delta^2}-R_{b,i} \\
(i = 1,2, ..., N-1 )$$

Drug to cyclodextrin complex varies according to

$$\frac {\partial C_{L-C}} {\partial \tau} = R_b \\
\frac{\partial C_{L-C}} {\partial \tau} = R_{b,i} \\
(i = 1,2, ..., N-1 )$$

With dimensionless boundary conditions at the center of the hydrogel given by

$$z = 0 , \frac {\partial C_L}{\partial z} = 0 \\
i = 0: \frac{\partial C_{L,0}}{\partial z} = 0 $$
For the release phase the dimensionless initial conditions are given by

$$ \tau = 0, C_L = \frac {C_L^{eq}} {C_0}, C_{L-C} = \frac {C_L^{eq}}{C_0} $$
which are the same initial conditions for the discretized version.

At the boundary of the hydrogel and drug release solution solution

$$\tau = 0, z = 1, C_L = 0 \\
\tau = 0, i = 1: C_{L,N}=0$$


$$\frac {dM_R} {d \tau} = - \frac {1}{2} \bigg[\frac{D} {k_2 \delta^2}\bigg] \frac {\partial C_l}{\partial z} \\
\frac {dM_R} {d \tau}=-\frac{1}{2}P_2\frac{C_{L,N}-C_{L,N-1}}{\Delta} $$

And with initial conditions given by:

$$M_R = 0, \tau = 0$$


```{r}
library(deSolve)
# Parameters
N   = 50 # Number of boxes
M_l = 2
C_o = 1 # Loading concentration after 


V_h = 0.00785 # hydrogel volume in ml
C_T = 15*(15/1000)/V_h # Total CD available
C_c = C_T  # Percent Free
k1  = 12.5*39.2 # M/h^-1
k2  = 39.2 # h^-1
K = k1/k2
L_eq = (C_o)/(1+K*C_c) # Ligand Equilibrium
complex_eq = C_o-((C_o)/(1+K*C_c))
thickness = 0.05
D   = 0.633*k2*(thickness^2) # Diffusion coefficient
delz = 1/N












# Parameters
# parameters <- c(
# p1 = k1*C_o/k2,
# p2 = D/(k2*delz^2),
# p3 = C/L_o

# Initial Value of State Variables

LIGAND = rep(L_eq/N, N)
LIGAND[N] <- 0
COMPLEX = rep(complex_eq/N, times = N)
RELEASE <- 0

state <- c(LIGAND = LIGAND,
           COMPLEX = COMPLEX,
           RELEASE = RELEASE)


affinity <- function(t, state , parms, ...){
  
p1 = k1*C_o/k2
p2 = D/(k2*thickness^2)
p3 = C_T/C_o

  LIGAND <- state[1:N]
  COMPLEX <- state[(N+1):(2*N)]
  RELEASE <- state[2*N+1]
  dLIGAND <- rep(0, times = N)
  dCOMPLEX <- rep(0, times = N)
  Rb <- p1*LIGAND*(p3 - COMPLEX) - COMPLEX
  dRELEASE <- 0


for(i in 1:(N-1)){
Rb[i] <- p1*LIGAND[i]*(p3 - COMPLEX[i]) - COMPLEX[i]  
    if(i == 1) {
dLIGAND[i] <- p2 * (LIGAND[2] - LIGAND[1])/(delz^2)- Rb[1]}
    if(i > 1 & i < (N-1)){
dLIGAND[i] <- p2*(LIGAND[i + 1]-2*LIGAND[i] + LIGAND[i-1])/(delz^2) - Rb[i]}
    if(i == (N-1)){
dLIGAND[i] <-  p2*(-2*LIGAND[i-1]+LIGAND[i-2])/(delz^2) - Rb[i-1]}
  

dCOMPLEX[i] <- Rb[i] 
dRELEASE <- -(1/2)*(LIGAND[N]-LIGAND[(N-1)])/delz
      

}

dudt <- rep(0, 2*N)

for(i in 1:N){
  dudt[i] <- dLIGAND[i]
  dudt[i+N] <- dCOMPLEX[i]
}
 dudt[(2*N)+1] <- dRELEASE
  ncall <<- ncall+1 
return(list(c(dudt))) 
             
  }
 
ncall <- 0
times <- seq(0, 1000, by = 0.5)

out <- ode(func = affinity,
           times = times, 
           y = state, parms = NULL)




```



```{r}
library(tidyverse)

tidy_out <- as.data.frame(out) %>%
  gather(., key = "species", value = "concentration", -time) %>%
  separate(col = species, into = c("species", "z"), sep = "(?<=[:alpha:])(?=[:digit:])") %>%
  mutate(z = as.numeric(z)) %>%
  mutate(time = time/k2) # time in hrs

tidy_out %>%
  filter(species == "LIGAND") %>%
  ggplot(., aes(x= time, y = z))+
  geom_raster(aes(fill = `concentration`), hjust = 0.5, vjust= 0.5, interpolate = F)
```


In this trial we will follow William Shchieser example in his book of analyzing
partial differential equations in R using the method of lines. In general Schieser approaches the solution of PDEs by making use of three main parts.  
    1. A main program providing  
       i) All the files and fx sourced into the program which contain  the derivative approximations 
       ii) The parameters explored  
       iii) The choice for the spatial derivative  
       iv) The level of output  
       v) The initial conditions  
       vi) The ode call to solve the system of equations  
    2. A PDE definition function which calls an external function to solve the spatial derivative. 
    3. An external function to resolve the timestep  
```{r}
tidy_out %>%
  # filter(first(time)) %>%
  filter(species == "LIGAND") %>%
  filter(time < 0.0002) %>%
  summarise(Total_LIGAND = sum(concentration*C_o))
```

```{r}
tidy_out %>%
  filter(species == "RELEASE") %>%
  mutate(release = cumsum(concentration)) %>%
  ggplot(., aes(x = time, y = release))+
  geom_point()
```

