---
title: "Affinity Release Model"
author: "Edgardo Rivera-Delgado"
date: '2018-03-01'
output:
  word_document: default
  html_notebook: default
csl: biomaterials.csl
bibliography: QSPR.bib
---

# Equations for Drug Release

Release from affinity polymer disks can be modeled as described by Andrew Fu. [@Fu:2011cj]. 


In the next set of equations the continuous equation will be given followed by the
discretized version used to solve it with the method of lines. 

The spatial derivative is discretized as given

$$ \Delta = \frac{1}{N}, z_{i} = i\Delta \\
(i = 0,1,2,...,N)$$

The rate of binding within the hydrogel is given by

$$ R_b = \bigg[\frac{k_1C_0} {k_2}\bigg] C_L\bigg(\bigg[\frac{C_T}{C_0}\bigg]-C_{L-C}\bigg)-C_{L-C} \\
R_{b,i}= P_1C_{L,i}\bigg(P_3-C_{L-C,i}\bigg)-C_{L-C,i}$$ 

Free drug in hydrogel varies according to

$$\frac {\partial C_L}{\partial \tau} = \bigg[ \frac {D}{k_2 \delta^2} \bigg ]\frac{\partial^2C_L}{\partial z^2}-R_b  \\
\frac{dC_{L,i}}{d\tau} = P_2 \frac{C_{L,i+1}+2C_{L,i}+C_{L,i-1}}{\Delta^2}-R_{b,i} \\
(i = 1,2, ..., N-1 )$$


Drug to cyclodextrin complex varies according to

$$\frac {\partial C_{L-C}} {\partial \tau} = R_b \\
\frac{\partial C_{L-C}} {\partial \tau} = R_{b,i} \\
(i = 1,2, ..., N-1 )$$

With dimensionless boundary conditions at the center of the hydrogel given by

$$z = 0 , \frac {\partial C_L}{\partial z} = 0 \\
i = 0: \frac{\partial C_{L,0}}{\partial z} = 0 $$
For the release phase the dimensionless initial conditions are given by

$$ \tau = 0, C_L = \frac {C_L^{eq}} {C_0}, C_{L-C} = \frac {C_L^{eq}}{C_0} $$
which are the same initial conditions for the discretized version.

Conditions at the boundary of the hydrogel and drug release solution 

$$\tau = 0, z = 1, C_L = 0 \\
\tau = 0, i = 1: C_{L,N}=0$$


$$\frac {dM_R} {d \tau} = - \frac {1}{2} \bigg[\frac{D} {k_2 \delta^2}\bigg] \frac {\partial C_l}{\partial z} \\ 
\frac {dM_R} {d \tau}=-\frac{1}{2}P_2\frac{C_{L,N}-C_{L,N-1}}{\Delta}$$


$$ \frac {dM_R} {d \tau} = - \frac {1}{2} \bigg[P_2\bigg] \frac {\partial C_l}{\partial z}$$

And with initial conditions given by:

$$M_R = 0, \tau = 0$$

$$ P_1 = \bigg[\frac{K_1*C_0}{K_2}\bigg], P_2 = \bigg[\frac{D}{K_2\delta^2}\bigg], P_3 = \bigg[\frac{C_T}{C_0}\bigg]$$

This model translates to deSolve in the following way
```{r}
library(deSolve)

# Parameters
N   = 50 # Number of boxes (dimensionless)
M_l = .002 # Initial Loading concentration in millimoles.
bcd_MW = 1134.98 # (mg/millimol)
bcd_disk_mass = 15 # (mg)
V_h = 0.00785 # hydrogel volume in (ml)
k2  = 36.9 # h^-1
k1  = 303.7*k2*1000 # mM^-1*h^-1
K = k1/k2 # mM^-1

C_T = (0.7*bcd_disk_mass/bcd_MW)/V_h # mmol/ml == mM
# C_T = C_c + C_LC, which means that the total amount of cd is fixed and exists 
# in either the free of ligand (C_c) or bound (C_LC )state

C_o = M_l/V_h # Loading concentration  (mg/ml) C_o = C_LC_eq+C_L_eq = M_l/V_h. Should be below C_T

# Total CD free to bind to drug
# bound_fraction <- 
C_c = C_T - C_o     # All Drug Bound
# C_c = C_T - C_o * 0 # ALL CD FREE
# C_c = C_T - C_o * bound_fraction 
# C_c = C_o

L_eq = (C_o)/(1+K*C_c) # Ligand Equilibrium
complex_eq = C_o-((C_o)/(1+K*C_c)) # 
thickness = 0.05 # cm
D   = 0.933*k2*(thickness^2) # Diffusion coefficient
delz = 1/N # dimensionless



LIGAND = rep(L_eq/(C_o), N)
LIGAND[N] <- 0
COMPLEX = rep(complex_eq/(C_o), times = N)
RELEASE <- 0

state <- c(LIGAND = LIGAND,
           COMPLEX = COMPLEX,
           RELEASE = RELEASE)


affinity <- function(t, state , parms, ...){
  
p1 = k1*C_o/k2
p2 = D/(k2*thickness^2)
p3 = C_T/C_o

  LIGAND <- state[1:N]
  COMPLEX <- state[(N+1):(2*N)]
  RELEASE <- state[2*N+1]
  dLIGAND <- rep(0, times = N)
  dCOMPLEX <- rep(0, times = N)
  Rb <- p1*LIGAND*(p3 - COMPLEX) - COMPLEX
  dRELEASE <- 0


for(i in 1:(N-1)){
Rb[i] <- p1*LIGAND[i]*(p3 - COMPLEX[i]) - COMPLEX[i]  
    if(i == 1) {
dLIGAND[i] <- p2 * (LIGAND[2] - LIGAND[1])/(delz^2)- Rb[1]}
    if(i > 1 & i < (N-1)){
dLIGAND[i] <- p2*(LIGAND[i + 1]-2*LIGAND[i] + LIGAND[i-1])/(delz^2) - Rb[i]}
    if(i == (N-1)){
dLIGAND[i] <-  p2*(-2*LIGAND[i-1]+LIGAND[i-2])/(delz^2) - Rb[i-1]}
  

dCOMPLEX[i+1] <- Rb[i+1] 
dRELEASE <- -(1/2)*(LIGAND[N]-LIGAND[(N-1)])/delz
      

}

dudt <- rep(0, 2*N)

for(i in 1:N){
  dudt[i] <- dLIGAND[i]
  dudt[i+N] <- dCOMPLEX[i]
}
 dudt[(2*N)+1] <- dRELEASE
  ncall <<- ncall+1 
return(list(c(dudt))) 
             
  }
 
ncall <- 0
times <- seq(0, 1000, by = 1)

out <- ode(func = affinity,
           times = times, 
           y = state, parms = NULL)




```

Now checking the concentrations within the hydrogel below. We can plot the ligand
and complex concentration side by side. You can see that the concentration of 
complex (yellowish region) is much higher relative to the ligand which appears
pitch blue for the duration of the experiment. Does this means that the ligand
is instantaneously depleted from the matrix?

```{r}
library(tidyverse)

tidy_out <- as.data.frame(out) %>%
  gather(., key = "species", value = "concentration", -time) %>%
  separate(col = species, 
           into = c("species", "z"), 
           sep = "(?<=[:alpha:])(?=[:digit:])") %>%
  mutate(z = as.numeric(z)) %>%
  mutate(time = time/k2) # time in hrs

library(viridis)
library(ggthemes)
tidy_out %>%
  filter(species %in% c("LIGAND", "COMPLEX")) %>%
  ggplot(., aes(x= time, y = z))+
  geom_raster(aes(fill = `concentration`),
              hjust = 0.5,
              vjust= 0.5, 
              interpolate = F)+
  facet_wrap(~species)+
  scale_fill_viridis()+
  theme_tufte()
```

In reality the ligand is not depleted from the matrix but present in very low 
concentrations when compared to the complex. If we plot just the ligand 
concentration we get a different picture as shown below.
```{r}
tidy_out %>%
  filter(species %in% c("LIGAND")) %>%
  ggplot(., aes(x= time, y = z))+
  geom_raster(aes(fill = `concentration`),
              hjust = 0.5,
              vjust= 0.5, 
              interpolate = F)+
  facet_wrap(~species)+
  scale_fill_viridis()+
  theme_tufte()
```




```{r}
tidy_out %>%
  filter(species == "RELEASE") %>%
  mutate(release = 100*cumsum(concentration)*100) %>%
  ggplot(., aes(x = time, y = release))+
  geom_point()+
  theme_hc()
```

```{r}
tidy_out %>%
  filter(species == "COMPLEX") %>%
  mutate(bound = 100*sum(concentration)) %>%
  ggplot(., aes(x = time, y = bound))+
  geom_point()+
  theme_hc()
```

