---
title: "Reproducible example for syg dynamic modeling group"
date: "2019-01-09"
output: html_notebook
---

# Problem

I am trying to model a diffusion-reaction process based on a paper by Fu et.al 
but I seem not to demonstrate mass is conserved for the system as a test for my implementation.

The paper predicts the release of a small molecule out of the top and bottom faces of of a 
thin cylindrical polymer with no wall transport. Within the polymer the molecule 
can either diffuse or it can react reversibly with a host. The resulting system of 
ode is solved with the method of lines. Since cylinders are symmetricals 

My goal is to understand the amount of (1) bound drug and (2) free drug within the 
polymer matrix at any given time. Most importantly, the modeling is used to simulate the 
cumulative release process out of the polymer based on the paraemters. The paper 
implements the solution for the system of equations in matlab and I would like to 
translate this approach to R using deSolve. I am having the following problems

I wanted help answering
(1) if the code below seems like a reasonable implementation of the model
in deSolve
(2) Do you have any ideas why I may be experiencing a lack of mass conservation in the system? Wrong calculation or logic?
(3) Any general suggestions or feedback is always appreciate

# The ode function
```{r}
affinity <- function(t, state , parms, ...){


# Extracting values passed onto the function through the state argument
  LIGAND     <- state[1:N]
  COMPLEX    <- state[(N+1):(2*N)]
  RELEASE    <- state[2*N+1]

# Initializing the derivatives vectors
  dLIGAND    <- rep(0, times = N)
  dCOMPLEX   <- rep(0, times = N)
  dRELEASE   <- 0

  
# Method of Lines  
Rb           <- p1 * LIGAND * (p3 - COMPLEX) - COMPLEX

# Calculating the rate of change of the state variables

## This layer represents no transport in the bottom face 
dLIGAND[1]   <- p2 * (LIGAND[2] - LIGAND[1]) / (delz^2) - Rb[1]
## This layer is the polymer to solution interface
dLIGAND[N-1] <- p2 * (-2*LIGAND[N-1] + LIGAND[N-2]) / (delz^2) - Rb[N-1]
## These are the calculations for the layers in between top and center
for(i in 2:(N-2)){
dLIGAND[i]   <- p2 * (LIGAND[i + 1] - 2*LIGAND[i] + LIGAND[i-1]) / (delz^2) - Rb[i]
}

dCOMPLEX     <- Rb

dRELEASE     <- -(1/2)*p2*(LIGAND[N]-LIGAND[(N-1)])/(delz)


dudt <- c(dLIGAND, dCOMPLEX, dRELEASE)

ncall <<- ncall+1 

return(list(c(dudt))) 
             
  }
```



# The parameters and state variables
```{r}

# Parameters
N   =    50.0         # Number of boxes 
M_l =     0.000426    # Initial drug in gel  in millimoles.
k2  =    36.9         # h^-1
k1  = 11206.53        # h/M
V_h =     0.00785     # mL
K   = k1/(1000*k2)    # mM^-1
C_o = M_l/V_h         # Loading concentration mM
C_T = 0.008810573/V_h # 0.1178505 mM
C_c = 0.000001*C_T    # Free cyclodextrin pockets

L_eq = ((C_o) / (1 + K * C_c))               # Ligand Equilibrium
complex_eq = (C_o - ((C_o) / (1 + K * C_c))) # 

delz = 1/N # dimensionless

# Dimensionless parameters
p1 = k1 * C_o / k2
p2 = 0.933       # From Paper
p3 = C_T/C_o

LIGAND <- rep(L_eq/(C_o*(N-1)), times = N)
LIGAND[N] <- 0

COMPLEX <- rep(complex_eq/(C_o*N), times = N)

RELEASE <- 0

state <- c(LIGAND = LIGAND,
           COMPLEX = COMPLEX,
           RELEASE = RELEASE)

```

# Solving the equation
```{r}
library(deSolve)
ncall <- 0
times <- seq(0, 350 * k2, by = 1)
out <- ode(func = affinity,
           times = times, 
           y = state)
```

# Plotting

```{r}
library(tidyverse)

# Turning the desolve object to a tidy dataframe
tidy_out <- as.data.frame(out) %>% 
  select(-RELEASE) %>%
  gather(., key = "species", value = "concentration", -time) %>%
  separate(col = species, 
           into = c("species", "z"), 
           sep = "(?<=[:alpha:])(?=[:digit:])") %>%
  mutate(z = as.numeric(z)) %>%
  mutate(time = time/k2) # time in hrs

# Grabbing the Release variable. I expect it to go up
tidy_release_media <- out %>%
  as.data.frame() %>%
  select(time, RELEASE) %>%
  mutate(time = time / k2) 
  
# Grouping by time to get a sum of all ligand inside the polymer through time. I expect it to go down
tidy_release_polymer <- tidy_out %>%
  filter(species %in% c("LIGAND", "COMPLEX")) %>%
  group_by(time) %>% 
  summarise(concentration = sum(concentration))

# When I sum both I expect it to be a flat line due to the principle of mass conservation
tidy_release_polymer %>%
  inner_join(tidy_release_media, by = "time") %>%
  rename(release_polymer = concentration,
         release_media = RELEASE) %>%
  mutate(mass_conservation = release_polymer + release_media) %>%
  ggplot(aes(x = time, y = mass_conservation))+
  geom_point()

```


