---
title: "Diffusion-reaction model out of polymer cylinder"
author: "Edgardo Rivera-Delgado"
output:
  html_document:
    df_print: paged
date: '2019-01-09'
---

# Problem

I am trying to model a diffusion-reaction transport process based on a paper by 
[Fu et.al](https://link.springer.com/article/10.1007%2Fs10439-011-0336-z)
but fail to demonstrate mass is conserved for the system as a test of my
implementation.

The paper simulates the release of a small molecule out of the top and bottom 
faces of a thin cylindrical polymer with no wall transport. Within the polymer
cylinder, the molecule drug can either diffuse (LIGAND) or it can react
reversibly  with a non-diffusing host (COMPLEX). The resulting system of ode is
solved with the method of lines. Since cylinders are symmetricals the system of 
equations is simplified to have no release from the bottom face and transport 
occurs only through the top face.

The model is used to simulate the cumulative release process of a drug out of 
the polymer cylinder into a liquid media. The paper by Fu implements the 
solution in matlab using ode15s and I would like to translate this approach to R 
using deSolve. I wanted to ask for your help answering the following questions:

(1) Does the code below seems like a reasonable implementation of the model
in deSolve?
(2) Do you have any ideas why I may be experiencing a lack of mass conservation 
in the system? Wrong code, wrong implementation or other reason?

---
Any general suggestions or feedback is always appreciated. I wanted to be as 
clear as possible so ended up with a verbose post. In the future, I can reduce 
the length if that's preferred by the list members and/or moderator. Thank you.

# The ode function
```{r}
affinity <- function(t, state , parms, ...){

# LIGAND + host <-> COMPLEX (LIGAND:host)
  
# The LIGAND is free to diffuse along the vertical length (N) of the polymer 
# cylinder. The host remains constant as it is part of the polymer cylinder 
# matrix. The LIGAND can  reversibly bind to the the host to form a COMPLEX. 
# Only the free LIGAND diffuses into the liquid media and is called the RELEASE.


# Extracting values passed into the function through the state argument
  LIGAND     <- state[1:N]
  COMPLEX    <- state[(N+1):(2*N)]
  RELEASE    <- state[2*N+1]

# Initializing the derivatives vectors
  # dLIGAND    <- rep(0, times = N)
  # dCOMPLEX   <- rep(0, times = N)
  # dRELEASE   <- 0

  
# Method of Lines  
## Binding process to host
  for(i in 1:(N-1)){
  Rb[i]         <- p1 * LIGAND[i] * (p3 - COMPLEX[i]) - COMPLEX[i] 
}
## Diffusion process of the ligand 
  
  dLIGAND[1]   <- p2 * (LIGAND[2] - LIGAND[1]) / (delta^2) - Rb[1]
  
## This layer represents the polymer cylinder to liquid media interface
  
  dLIGAND[N-1] <- p2 * (-2 * LIGAND[N-1] + LIGAND[N-2]) / (delta^2) - Rb[N-1]
  
## These are the calculations for the layers in between top and center
  
for(i in 2:(N-2)){
  dLIGAND[i] <- 
    p2 * (LIGAND[i + 1] - 2 * LIGAND[i] + LIGAND[i-1]) / (delta^2) - Rb[i]
}
  
## Change in COMPLEX through the polymer cylinder
  
  dCOMPLEX     <- Rb

## LIGAND exiting into the liquid environment
  dRELEASE     <- -(1/2) * p2 * (LIGAND[N] - LIGAND[(N-1)]) / (delta) 

  dudt         <- c(dLIGAND, dCOMPLEX, dRELEASE)

ncall <<- ncall + 1 

return(list(dudt)) 
             
  }
```

# The parameters and state variables
```{r}

# Parameters

## Assigned

time  =   350.0         # Length of simulation in hrs

M_l   =     0.0004      # Initial drug in polymer cylinder in millimoles.
V_h   =     0.0785      # Volume of polymer cylinder mL

k2    =    36.9         # LIGAND decoupling rate from host in 1/hrs
k1    =    11.2     # LIGAND coupling rate to host in 1/mM*hrs
K     =   k1 / k2       # Binding strength between drug and host in mM^-1

N     =    50.0         # Number of boxes 
delta =    1 / N        # Spacing


## Estimated parameters

C_o = M_l / V_h                   # LIGAND in polymer cylinder in mM
C_T = 0.00881 / V_h           # Host concentration in polymer cylinder in mM 
C_c = 0.00001 * C_T               # Fraction of hosts without bound drug

## Equilibrium concentrations
### LIGAND  + COMPLEX (LIGAND:host) = C_o = Total LIGAND in polymer cylinder

ligand_eq  =        (C_o) / (1 + K * C_c) # Free LIGAND at time = 0 in mM
complex_eq = C_o -  (C_o) / (1 + K * C_c) # COMPLEX at time = 0  in mM

## Dimensionless parameters

p1           = k1 * C_o / k2
p2           = 0.933            
p3           = C_T / C_o

ligand_init  =  ligand_eq / C_o  # Dimensionless free LIGAND
complex_init = complex_eq / C_o  # Dimensionless COMPLEX
tau          =       time * k2   # Dimensionless time

# State variables 

### Free LIGAND divided into the total number of layers with the layer in direct 
### contact to the liquid media set to 0

LIGAND    <- c(rep((ligand_init / (N-1)), times = N-1), 0)  

### COMPLEX divided into the total number of layers

COMPLEX   <- rep((complex_init / N), times = N)   

### Initially there's no drug in the liquid environment

RELEASE   <- 0  

state     <- c(LIGAND = LIGAND,
               COMPLEX = COMPLEX,
               RELEASE = RELEASE)

```

# Solving the equation
```{r}
library(deSolve)
ncall <- 0

times <- seq(0, tau, by = 1)
desolve_output <- ode(func = affinity, 
                      times = times, 
                      y = state)
```

# Plotting

```{r}
library(tidyverse)


# At each time sum the concentration of free LIGAND and bound LIGAND (complex) 
# within the polymer cylinder to calculate the LIGAND depletion rate. I expect
# it to go down to 0.

tidy_release_polymer  <- as.data.frame(desolve_output) %>% 
  select(-RELEASE) %>%
  gather(., key = "species", value = "concentration", -time) %>%
  separate(col = species, 
           into = c("species", "z"), 
           sep = "(?<=[:alpha:])(?=[:digit:])") %>% # Separating the state 
                                                    # variable names from the
                                                    # vertical coordinates of 
                                                    # the polymer cylinder
  mutate(z = as.numeric(z)) %>%
  mutate(time = time / k2) %>% # Reverting back time from dimensionless to hrs
  filter(species %in% c("LIGAND", "COMPLEX")) %>% # 
  group_by(time) %>% 
  summarise(ligand_in_polymer = sum(concentration)) 

# Grabbing the RELEASE variable. I expect it to go up to 1.

tidy_release_media <- desolve_output %>%
  as.data.frame() %>%
  select(time, RELEASE) %>%
  mutate(time = time / k2) %>%
  rename(ligand_in_media = RELEASE) 
  
# When I sum both the drug remaining in the polymer cylinder and the drug 
# entering the solution I expect a flat line around 1 (+/- numerical error) due 
# to the principle of mass conservation. However, I obtain a depletion curve.

tidy_release_polymer %>%
  inner_join(tidy_release_media, by = "time") %>%
  mutate(mass_conservation = ligand_in_polymer + 100 * ligand_in_media) %>%
  ggplot(aes(x = time, y = mass_conservation)) +
  geom_point()

```



