---
title: "Diffusion-reaction model out of polymer cylinder"
author: "Edgardo Rivera-Delgado"
output:
  html_document:
    df_print: paged
date: '2019-01-09'
---

# Problem

I am trying to model a diffusion-reaction transport process based on a paper by 
[Fu et.al](https://link.springer.com/article/10.1007%2Fs10439-011-0336-z)
but fail to demonstrate mass is conserved for the system as a test of my implementation.

The paper simulates the release of a small molecule out of the top and bottom faces of a 
thin cylindrical polymer with no wall transport. Within the polymer cylinder, 
the molecule drug can either diffuse (LIGAND) or it can react reversibly  
with a non-diffusing host (COMPLEX). The resulting system of ode is solved with the method 
of lines. Since cylinders are symmetricals the system of equations is simplified 
to have no release from the bottom face and transport occurs only through the 
top face.

The model is used to simulate the cumulative release process out of the polymer cylinder. 
The paper by Fu implements the solution in matlab using ode15s and I would like to 
translate this approach to R using deSolve. I wanted to ask for your help answering
the following questions:

(1) Does the code below seems like a reasonable implementation of the model
in deSolve?
(2) Do you have any ideas why I may be experiencing a lack of mass conservation 
in the system? Wrong code, wrong implementation or other reason?

Any general suggestions or feedback is always appreciated. Thank you.

# The ode function
```{r}
affinity <- function(t, state , parms, ...){

# LIGAND + HOST <-> COMPLEX (LIGAND:HOST)
  
# The LIGAND is free to diffuse and the host is part of the polymer matrix. 
# In the COMPLEX form the LIGAND is said to be bound
# In the 

# Extracting values passed onto the function through the state argument
  LIGAND     <- state[1:N]
  COMPLEX    <- state[(N+1):(2*N)]
  RELEASE    <- state[2*N+1]

# Initializing the derivatives vectors
  dLIGAND    <- rep(0, times = N)
  dCOMPLEX   <- rep(0, times = N)
  dRELEASE   <- 0

  
# Method of Lines  
  Rb         <- p1 * LIGAND * (p3 - COMPLEX) - COMPLEX

# Calculating the rate of change of the state variables

## This layer represents no transport in the bottom face 
  
  dLIGAND[1]   <- p2 * (LIGAND[2] - LIGAND[1]) / (delta^2) - Rb[1]
  
## This layer represents the polymer to solution interface
  
  dLIGAND[N-1] <- p2 * (-2 * LIGAND[N-1] + LIGAND[N-2]) / (delta^2) - Rb[N-1]
  
## These are the calculations for the layers in between top and center
  
for(i in 2:(N-2)){
  dLIGAND[i]   <- p2 * (LIGAND[i + 1] - 2 * LIGAND[i] + LIGAND[i-1]) / (delta^2) - Rb[i]
}
  
## Change in bound drug through the polymer cylinder
  
  dCOMPLEX     <- Rb

  dRELEASE     <- -(1/2) * p2 * (LIGAND[N] - LIGAND[(N-1)]) / (delta) 


  dudt         <- c(dLIGAND, dCOMPLEX, dRELEASE)

ncall <<- ncall + 1 

return(list(c(dudt))) 
             
  }
```



# The parameters and state variables
```{r}

# Parameters

## Assigned

time  =   350.0         # Length of simulation in hrs

M_l   =     0.000426    # Initial drug in cylinder in millimoles.
V_h   =     0.0785     # Volume of polymer cylinder mL

k2    =    36.9         # LIGAND decoupling rate from host in 1/hrs
k1    =    11.20653     # LIGAND coupling rate to host in 1/mM*hrs
K     =   k1 / k2       # Binding strength between drug and host in mM^-1

N     =    50.0         # Number of boxes 
delta =    1 / N        # Spacing


## Estimated parameters

C_o = M_l / V_h         # Initial total drug concentration inside of the polymer in mM
C_T = 0.008810573 / V_h # Host concentration inside polymer cylinder in mM (*Bound to polymer, constant through time)
C_c = 0.001 * C_T    # Fraction of hosts without bound drug

## Equilibrium concentrations

ligand_eq  =        (C_o) / (1 + K * C_c) # Drug concentration not bound to host at time = 0 in mM
complex_eq = C_o -  (C_o) / (1 + K * C_c) # Drug concentration bound to host at time = 0  in mM

## Dimensionless parameters obtained from paper

p1           = k1 * C_o / k2
p2           = 0.933            
p3           = C_T / C_o

ligand_init  =  ligand_eq / C_o  # Dimensionless free drug
complex_init = complex_eq / C_o  # Dimensionless bound drug
tau          =       time * k2   # Dimensionless time

# State variables 

LIGAND    <- c(rep((ligand_init / (N-1)), times = N-1), 0)  # Free drug divided into the total number of layers, layer in direct contact to media is set to 0

COMPLEX   <- rep((complex_init / N), times = N)             # Bound drug divided into the total number of layers

RELEASE   <- 0                                              # Initial condition

state     <- c(LIGAND = LIGAND,
               COMPLEX = COMPLEX,
               RELEASE = RELEASE)

```

# Solving the equation
```{r}
library(deSolve)
ncall <- 0

times <- seq(0, tau, by = 1)
desolve_output <- ode(func = affinity, 
                      times = times, 
                      y = state)
```

# Plotting

```{r}
library(tidyverse)


# At each time sum the concentration of free drug (ligand) and bound drug (complex) within the 
# polymer to calculate the drug depletion rate. I expect this to go down to 0.

tidy_release_polymer  <- as.data.frame(desolve_output) %>% 
  select(-RELEASE) %>%
  gather(., key = "species", value = "concentration", -time) %>%
  separate(col = species, 
           into = c("species", "z"), 
           sep = "(?<=[:alpha:])(?=[:digit:])") %>% # Separating the column names from the vertical coordinates of the cylinder
  mutate(z = as.numeric(z)) %>%
  mutate(time = time / k2) %>% # Reverting back time from dimensionless to hrs
  filter(species %in% c("LIGAND", "COMPLEX")) %>% # 
  group_by(time) %>% 
  summarise(drug_in_polymer = sum(concentration)) # Calculate the sum of free ligand and bound

# Grabbing the RELEASE variable. I expect it to go up to 1

tidy_release_media <- desolve_output %>%
  as.data.frame() %>%
  select(time, RELEASE) %>%
  mutate(time = time / k2) %>%
  rename(drug_in_media = RELEASE) # Possibly multiply by 2*delta. But why?
  
# When I sum both the drug remaining in the polymer and the drug entering the solution
# I expect a flat line around 1 (+/- numerical error) due to the principle 
# of mass conservation.However, I obtain a depletion curve.

tidy_release_polymer %>%
  inner_join(tidy_release_media, by = "time") %>%
  mutate(mass_conservation = drug_in_polymer + drug_in_media) %>%
  ggplot(aes(x = time, y = mass_conservation)) +
  geom_point()

```



